Cooperation Provably Prevails in a Version of the Noisy Iterated Prisoner's Dilemma

### 2020-05-16 paper/paper.tex
change terminology
add tikz figures

\documentclass[11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{todonotes}
\usepackage{cleveref}

@@

\begin{definition}
  The \textit{prisoner's dilemma} is a symmetric two-player game with two actions, cooperate ($C$) and defect ($D$), where, if player 1 selects action $a$ and player 2 selects action $b$, player 1 gets reward
  \begin{equation*}
    r(a,b) = \begin{cases}
      R &\text{if $a = C, b = C$} \\
      T &\text{if $a = D, b = C$} \\
      S &\text{if $a = C, b = D$} \\
      P &\text{if $a = D, b = D$}
    \end{cases}
  \end{equation*}
  We require $T > R > P > S$ and $2R > T + S$. 
\end{definition}

@@

Notation-wise, we will use $c$ to denote states in $s$, $G_s(c)$ to denote the output at state $c$, and $T_s(c, a)$ to denote the state that $c$ transitions to upon receiving input $a$. For simplicity, we will also define the $\lnot$ operator such that $\lnot C = D$ and $\lnot D = C$, and $c_{\text{start}}(s)$ to be the start state of $s$.

@@

That is, when $s_1$ plays against $s_2$, we define its payoff to be the average payoff over all possible infinite sequences of moves. Note that the expectation is taken over the infinite sequence $(X_0, X_1, \ldots)$. The limit inside is thus simply a normal time-average limit of bounded real numbers, which clearly exists.

@@

\begin{definition}
  \label{timeaveragedistribution}
  The \textit{time average distribution} of the $s_1$-$s_2$ Markov chain given the start state $(a, b)$, denoted $\pi^{(a, b)}$, is the distribution such that \begin{equation*}
    \pi_{c_1, c_2}^{(a, b)} = E \left[ \text{fraction of time in state $(c_1,c_2)$} \mid \text{initial state is } (a, b)  \right]
  \end{equation*}
  where the fraction of time is taken over the infinite sequence $(X_0, X_1, \ldots)$.
\end{definition}

\begin{lemma}
  \label{payofftimeaverage}
  The payoff when $s_1$ plays against $s_2$ is
  \begin{equation*}
    v_{s_1}(s_2) = \sum_{(c_1,c_2) \in S_{s_1,s_2}} \pi_{c_1,c_2}^{(c_{\text{start}}(s_1), c_{\text{start}}(s_2))} \cdot r(G_{s_1}(c_1),G_{s_2}(c_2)).
  \end{equation*}
\end{lemma}

We may also make $r(G_{s_1}(c_1),G_{s_2}(c_2))$ into a vector, denoted by $r$, and write this as \begin{equation*}
  v_{s_1}(s_2) = \pi^{(c_{\text{start}}(s_1), c_{\text{start}}(s_2))} \cdot r.
\end{equation*}

@@

\begin{definition}
  A strategy $s_1$ is \textit{evolutionarily stable} if there exists parameters $p_0$ and $\alpha$, both in $(0,1)$, such that for all $p < p_0$, and all $\epsilon < \alpha$, $s_1$ is not $\epsilon$-invadable.
\end{definition}

@@

\begin{figure}
  \label{tftfigure}
  \includegraphics[width=4cm]{tft.jpg}
  \centering
  \caption{TFT.}
\end{figure}
\begin{figure}
  \label{pavlovfigure}
  \includegraphics[width=4cm]{pavlov.jpg}
  \centering
  \caption{Pavlov.}
\end{figure}
  \todo{use tikz for these figures!}

@@

      To do that, we want to prove that for all $p_0, \alpha \in (0,1)$, there exists $p < p_0$ and $\epsilon < \alpha$, such that $s_1$ is $\epsilon$-invadable. We choose $\epsilon = \alpha / 2$, and present a strategy $s_2$ that can invade $s_1$ for sufficiently small $p$.

      We create the strategy $s_2$ as follows. First, copy the entire $s_1$ machine into $s_2$. Suppose that the state corresponding to the start state of $s_1$ is $c_s$. Recall that the output at $c_s$ is $G(c_s)$, and that the state $s$ goes to upon perceiving the opponent move $G(c_s)$ is $T(c_s, G(c_s))$. Now, create two new states: $c_0$ and $c_1$. Define the transitions as \begin{align*}
        T(c_0, G(c_s)) &= T(c_s, G(c_s))\\
        T(c_0, \lnot G(c_s)) &= c_1 \\
        T(c_1, \cdot) &= c_1
      \end{align*}
      and the outputs as \begin{align*}
        G(c_0) &= \lnot G(c_2)\\
        G(c_1) &= C.
      \end{align*}
      Let the start state of $s_2$ be $c_0$. \todo{add a figure!! the construction is simple but this description is deceivingly hard}

@@

    \section{Appendix: Non-stationary limiting distributions}

>>>>

\documentclass[11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{todonotes}
\usepackage{cleveref}
\usepackage{tikz}
\usetikzlibrary{backgrounds,arrows,automata,positioning}

@@

\begin{definition}
  The \textit{prisoner's dilemma} is a symmetric two-player game with two actions, cooperate ($C$) and defect ($D$), where, if player 1 selects action $a$ and player 2 selects action $b$, player 1 gets the reward
  \begin{equation*}
    r(a,b) = \begin{cases}
      R &\text{if $a = C, b = C$} \\
      T &\text{if $a = D, b = C$} \\
      S &\text{if $a = C, b = D$} \\
      P &\text{if $a = D, b = D$}
    \end{cases}
  \end{equation*}
  We require $T > R > P > S$ and $2R > T + S$. 
\end{definition}

@@

Notation-wise, we will use $c$ to denote states in $s$, $G_s(c)$ to denote the output at state $c$, and $T_s(c, a)$ to denote the state that $c$ transitions to upon receiving input $a$. For simplicity, we will also define the $\bar{\;}$ operator such that $\bar{C} = D$ and $\bar{D} = C$, and $c_{\text{start}}(s)$ to be the start state of $s$.

@@

That is, when $s_1$ plays against $s_2$, we define its payoff to be the average payoff over all possible infinite sequences of moves. Note that the expectation is taken over the infinite sequence $(X_0, X_1, \ldots)$. The limit inside is thus simply a normal time average limit of bounded real numbers, which clearly exists.

@@

\begin{definition}
  \label{timeaveragedistribution}
  The \textit{time average distribution} of the $s_1$-$s_2$ Markov chain given the start state $(a, b)$, denoted $\pi^{(a, b)}$, is the distribution such that \begin{equation*}
    \pi_{c_1, c_2}^{(a, b)} = E \left[ \text{fraction of time in state $(c_1,c_2)$} \mid \text{initial state is } (a, b)  \right]
  \end{equation*}
  where the expected fraction of time is taken over the infinite sequence $(X_0, X_1, \ldots)$.
\end{definition}

We will use $\pi$ to refer to $\pi^{(c_{\text{start}}(s_1), c_\text{start}(s_2))}$.

\begin{lemma}
  \label{payofftimeaverage}
  The payoff when $s_1$ plays against $s_2$ is
  \begin{equation*}
    v_{s_1}(s_2) = \sum_{(c_1,c_2) \in S_{s_1,s_2}} \pi_{c_1,c_2} \cdot r(G_{s_1}(c_1),G_{s_2}(c_2)).
  \end{equation*}
\end{lemma}

We may also make $r(G_{s_1}(c_1),G_{s_2}(c_2))$ into a vector, denoted by $r$, and write this as the dot product \begin{equation*}
  v_{s_1}(s_2) = \pi \cdot r.
\end{equation*}

@@

\begin{definition}
  A strategy $s_1$ is \textit{evolutionarily stable} if there exists parameters $p_0$ and $\epsilon_0$, both in $(0,1)$, such that for all $p < p_0$, and all $\epsilon < \epsilon_0$, $s_1$ is not $\epsilon$-invadable.
\end{definition}

@@

\begin{figure}
  \label{pavlovfigure}
\begin{tikzpicture}[node distance=2cm]
  \node[state,initial] (s_0) {$C$};
  \node[state] (s_1) [right of=s_0] {$D$};

  \path[->] (s_0) edge [loop above] node {$C$} ();
  \path[->] (s_0) edge [bend left] node [above] {$D$} (s_1);
  \path[->] (s_1) edge [loop above] node {$C$} ();
  \path[->] (s_1) edge [bend left] node [below] {$D$} (s_0);
\end{tikzpicture}
  \centering
  \caption{Pavlov.}
\end{figure}

\begin{figure}
  \label{tftfigure}
\begin{tikzpicture}[node distance=2cm]
  \node[state,initial] (s_0) {$C$};
  \node[state] (s_1) [right of=s_0] {$D$};

  \path[->] (s_0) edge [loop above] node {$C$} ();
  \path[->] (s_0) edge [bend left] node [above] {$D$} (s_1);
  \path[->] (s_1) edge [loop above] node {$D$} ();
  \path[->] (s_1) edge [bend left] node [below] {$C$} (s_0);
\end{tikzpicture}
  \centering
  \caption{Tit for tat.}
\end{figure}
  % \begin{tikzpicture}[scale=0.8, show background rectangle]
  %   \draw (0,0) -- (2,1);
  %   \draw (1,1) circle (3);
  %   \path[draw] (0,0) -- (1,1) -- (2,0);
  % \end{tikzpicture}

@@

      To do that, we want to prove that for all $p_0, \epsilon_0 \in (0,1)$, there exists $p < p_0$ and $\epsilon < \epsilon_0$, such that $s_1$ is $\epsilon$-invadable. We choose $\epsilon = \epsilon_0 / 2$, and present a strategy $s_2$ that can invade $s_1$ for sufficiently small $p$.

      We create the strategy $s_2$ as follows. First, copy the entire $s_1$ machine into $s_2$. Suppose that the state corresponding to the start state of $s_1$ is $c$, and that the output at $c$ is $\alpha$, and that the state $s$ goes to upon perceiving the opponent move $\alpha$ is $c' = T(c, \alpha)$. Now, create two new states: $c_0$ and $c_1$. Define the transitions as \begin{align*}
        T(c_0, \alpha) &= c'\\
        T(c_0, \bar{\alpha}) &= c_1 \\
        T(c_1, \cdot) &= c_1
      \end{align*}
      and the outputs as \begin{align*}
        G(c_0) &= \lnot G(c_2)\\
        G(c_1) &= C.
      \end{align*}
      Let the start state of $s_2$ be $c_0$.

      \begin{figure}
        \centering
        \begin{tikzpicture}[node distance=2cm]
          \node[state, initial] (c_0) {$\bar{\alpha}$};
          \node[state] (c_1) [below right of=c_0] {$C$};
          \node[state] (c_prime) [above right of=c_0] {};
          \node[state] (c) [above left of=c_prime] {$\alpha$};

          \path[->] (c_0) edge node [above right] {$\bar{\alpha}$} (c_1);
          \path[->] (c_1) edge [loop right] node [right] {$C,D$} ();
          \path[->] (c_0) edge node [above left] {$\alpha$} (c_prime);
          \path[->] (c) edge node [below left] {$\alpha$} (c_prime);
        \end{tikzpicture}
        \caption{Constructon of invasion strategy, used in the proof of \cref{evolutionarystable1}}
        \label{invasionstrategy}
      \end{figure}

@@

    \section{Appendix: Time average distributions}
